---
title: "Statistical Modeling of Animal Movement with R Package smam"
fontsize: 12pt
documentclass: article
geometry:
  margin = 1in
bibliography: blog.bib
biblio-style: asa
output: pdf_document
---


Among the many R packages for animal movement [@Joo:etal:2020],
`smam` [@Rpkg:smam] is unique in that it allows an animal to stay still
instead of perpetual movement.
This model is known as the moving-resting process, in which the
holding times in the moving stage and the resting stage are
exponentially distributed, and the moving stage is modeled by a
Brownian motion. The model was first proposed in 2014 [@Yan:etal:2014],
but the measurement errors from the telemetric or GPS
devices are not satisfactorily handled until our recent paper in
Methods in Ecology and Evolution [@hu:etal:2021], which presents a
solid and practical inference process for the
moving-resting model with measurement errors (MRME).


# Fitting the MRME model to animal movement data


As an illustration, consider the movement data of a female mountain lion
with id `f109`  in the Gros Ventre Mountain area, Wyoming. This data set was
analyzed in @hu:etal:2021.  First, we need to load the package and GPS
location data `f109raw`.  The datasets analyzed
by this package should be formatted as a `data.frame` whose first column is the
observation time and the other columns are location coordinates.
In `f109`, column `t` is the time of records collected in seconds, and
columns `dE` and `dN` are the centered Universal Transverse Mercator
(UTM) easting and northing in meters. The trace of this animal
is shown in the following plot. The distance should be expressed in 
kilometers before analysis and plotting.
```{r, echo = TRUE}
library(smam)
head(f109raw)

## converting the distance units to kilometers
f109raw <- within(f109raw, {dE = dE / 1000; dN = dN / 1000})
```

```{r, echo = TRUE, eval = TRUE, fig.width = 8}
with(f109raw, plot(t, dE, type = 'l', ylim = c(-20, 30),
                   xlab = "time (year)", ylab = "location (km)"))
with(f109raw, lines(t, dN, lty = 2))
legend(2009.2, 30, c("easting", "northing"), lty = c(1, 2))
```

The mountain lions behave differently in summer and winter seasons.
As a result, these two seasons have to be modeled differently. 
The summer is defined as June, July, and August.
The winter is defined as December, January, and February
We covert time units from years
to hours for ease of interpretation of the Brownian motion mobility parameter.

```{r, echo=TRUE}
## define season
f109raw$season <- with(f109raw, {
    tt  <- t %% 1
    ifelse(tt > 5/12 & tt <= 8/12, "summer",
             ifelse(tt > 11/12 | tt <= 2/12, "winter", "other"))
})

## define time in hours
f109raw$hr <- (f109raw$t - 2009) * 365.25 * 24
## define year with March as starting month
f109raw$year <- (f109raw$t - 2/12) %/% 1

## subset of 2012 summer
summer2012 <- subset(f109raw, season == "summer" & year == 2012,
                     select = c("hr", "dE", "dN"))

## subset of 2011-2012 winter
winter2011 <- subset(f109raw, season == "winter" & year == 2011,
                     select = c("hr", "dE", "dN"))

head(summer2012)
head(winter2011)
```

Now, the MRME model can be applied.
The estimation results are displayed in the following order:

+ $\lambda_M$, rate parameter of the exponential holding time of moving; 
+ $\lambda_R$, rate parameter of the exponential holding time of
  resting;
+ $\sigma$, mobility parameter of the Brownian motion in moving;
+ $\sigma_{\epsilon}$, standard error of the measurement error.


```{r, echo=TRUE, eval=FALSE, cache=TRUE}
## fit 2012 summer data to the MRME model
## This takes a while
fit_summer <- fitMRME(summer2012,
                      start = c(4, 0.4, 1, 0.05))
estimate(fit_summer)
#       lamM       lamR      sigma    sig_err 
# 2.84049099 0.17869261 1.33492477 0.01852022

## fit 2011 winter data to the MRME model
fit_winter <- fitMRME(winter2011,
                      start = c(4, 0.4, 1, 0.05))
estimate(fit_winter)
#        lamM        lamR       sigma     sig_err 
# 6.304029786 0.118513328 1.530265318 0.009068519
```

We use summer estimators as an example.
The estimates $\hat{\lambda_M}=2.84 hr^{-1}$ and
$\hat{\lambda_R}=0.18 hr^{-1}$
mean that the average durations of moving and
resting are $1/2.84=0.35$ and $1/0.18=5.56$ hours.
The estimate $\hat{\sigma}=1.33 km/hr^{1/2}$
tells us that if the lion keeps moving without stopping 
for one hour, the average departure from original
point is $1.33 km$. The last parameter estimate, 
$\hat{\sigma_{\epsilon}}=0.019 km$, is the standard error of the
measurement error.


The package provides standard errors for inferences based on Godambe
information and parametric bootstrap. The  
recommended method is a parametric bootstrap, where
`nBS` specifies the number 
of bootstraps, and `numThreads` specifies the number of threads to be
used for multicore computers. The following code
provides the estimated covariance matrix of estimators.
Taking square root of diagonal elements gives us  
the standard errors of the estimators.


```{r, echo=TRUE, eval=FALSE}
vcov(fit_summer, nBS = 100, numThreads = 6)
#  |=============================================| 100%
#               [,1]          [,2]          [,3]          [,4]
# [1,]  2.111367e-01  1.483937e-03  3.629777e-02 -6.775386e-05
# [2,]  1.483937e-03  1.963202e-04 -2.547285e-05 -2.106885e-06
# [3,]  3.629777e-02 -2.547285e-05  1.127234e-02 -1.014327e-05
# [4,] -6.775386e-05 -2.106885e-06 -1.014327e-05  7.556806e-07
```

Summer data over multiple years can be pooled to form a more precise
estimation. 

```{r, echo = TRUE}
## define summer batch variable
summer <- subset(f109raw, season == "summer",
                 select = c("hr", "dE", "dN", "year"))
```
Now, to fit the MRME model to only summer part of the dataset
one should proceed as follow. The additional argument `segment="year"`
identifies which observations are in the same year; the composite
loglikelihoods for different years are added up as the pooled
composite loglikelihood.

```{r, echo = TRUE, eval = FALSE}
fit <- fitMRME(summer, start = c(5, 0.5, 1, 0.01),
               segment = "year")
estimate(fit)
#       lamM       lamR      sigma    sig_err
# 3.82490120 0.16065612 1.86454989 0.02043802
```

# Simulation from for the MRME model

One might also want to generate movement data from the MRME model for
exploration or simulation purposes.
In the following example, we show how to generate
a two-dimensional dataset from MRME model by R function
`rMRME`. Here, we use `time` to indicates
time points at which observations are to be simulated.
Again, `lamM`, `lamR`, `sigma`, `sig_err` are four parameters of
MRME model that describe the duration of moving and 
resting, mobility parameter $\sigma$ and measurement
error. The initial state of this
realization is set as moving by `s0="m"` (and you can
also set it to resting by `s0="r"`). The dimension
of this generated data is 2 that is specified with
`dim = 2`. The columns `X1` and `X2` in
`dat` are the two-dimensional location coordinates.
The trajectories of this simulated data are shown
in the following plot.

```{r, echo = TRUE}
tgrid <- seq(0, 10*500, length=500)
set.seed(123)
dat <- rMRME(time = tgrid, lamM = 1, lamR = 0.5, sigma = 1,
             sig_err = 0.01, s0 = "m", dim = 2)
head(dat)
```

```{r, echo = FALSE, eval = TRUE}
with(dat, plot(time, X1, type = 'l', xlab = "time", ylab = "location", ylim = range(c(X1, X2))))
with(dat, lines(time, X2, lty = 2))
legend(10, 45, c("dimension 1", "dimension 2"),lty = c(1, 2))
```

# Other models implemented in `smam`

In addition to the MRME, there are four models implemented in the package.

- **Moving-resting (MR) process** [@Yan:etal:2014; @Pozd:etal:2017]:
    models animal movement with two
    states, moving and resting. The transition between two
    states is modeled by telegraph process.
    Resting is a motionless state and moving
    is modeled by Brownian motion. Functions `fitMR()` and `rMR()`
	provide fitting and simulation from the model, respectively.
- **Brownian motion with measurement error (BMME)** [@Pozd:etal:2014]:
    models animal movement
    by Brownian motion only with measurement error modelled by added
    Gaussian noise.  Functions `fitBMME()` and `rBMME()` provide
    fitting and simulation from the model, respectively.
- **Moving-resting-handling (MRH) process** [@Pozd:etal:2018]:
    models animal movement
    with three states: moving, resting, and handling. Moving state
    is modeled by Brownian motion. Both resting and handling states
    are motionless states with duration times modeled by
    two different exponential distributions. This model assumes animal always
    switches from a motionless state to the moving state. But when animal
    switches from the moving state, it may end with resting or handling.
    Functions `fitMRH()` and `rMRH()` provide fitting and simulation
    from the model, respectively.
- **Moving-moving (MM) process**[@hu:etal:2021]: 
    modified the moving-resting process by
    replacing the resting state with a second moving state 
    with a relatively lower mobility. These two
    moving states are assumed to follow two Brownian motions with
    different mobility parameters. The duration time of these two
    states are also described by exponential distributions with different
    rate parameters. Functions `fitMM()` and `rMM()` provide fitting
    and simulation from the model, respectively.

The `vcov` method has been implemented for each of the five models.
See the documentation within the package for more details.

# References
