---
title: "smam: Statistical Modeling for Animal Movements"
author: "Jun Yan, Vladimir Pozdnyakov, Chaoran Hu"
fontsize: 12pt
documentclass: article
margin: 1in
output: pdf_document
---

#1. Introduction

This package is designed for animal movement analysis. We totally implemented
three models, Brownian motion with measurement error [4], moving-resting process
with embedded Brownian motion [1,2], and moving-resting-hunting process with embedded
Brownian motion [3]. These three models can also be fitted with specific parts of
dataset, which is called *segment estimation* in this vignette. From the perspective
of ecology and animal movement, segment estimation can be hired for seasonal analysis.
We also provided toolbox for seasonal analysis within this package, which can subset
data according to given date interval.

In this vignette, we play with the GPS dataset of a mountain lion and quantify her
movement behavior. First of all, let us take a look at the dataset.

```{r, include=FALSE}
load("f109.RData")
```

```{r, echo=TRUE}
library(smam)
head(f109)
```

The first column, 'date', is the date when the observation is collected. The second column,
'time', is the corresponding time line of observations. The last two columns, 'centerE' and
'centerN', are the GPS coordinates.

#2. Whole dataset analysis

In this section, we are going to show how to use whole dataset to fit BMME, moving-resting
model and moving-resting-handling model with R function `fitBMME`, `fitMR` and `fitMRH`.
Let's see the standard format of dataset used by these functions first. It should be a
`data.frame` whose first column is the observation time and the other columns are the
location coordinates, as showing below.

```{r, echo = TRUE}
dat1 <- f109[, -1]
head(dat1)
```

Now, we can fit the above data by the following code. (We do not show the result,
because the estimation process for moving-resting-handling model is
time consuming and maybe take several hours.)

```{r, eval=FALSE, echo=TRUE}
## fit BMME model
fitBMME(dat1, start = c(0.5, 0.5))

## fit moving-resting model with full likelihood
fitMR(dat1, start = c(10, 0.1, 1), likelihood = "full")

## fit moving-resting model with composite likelihood
fitMR(dat1, start = c(10, 0.1, 1), likelihood = "composite")

## fit moving-resting-handling model (serial computation)
## numThread is less or equal to 1 will call serial computation
fitMRH(dat1, start = c(4, 0.4, 0.1, 1, 0.5), numThreads = 1)

## fit moving-resting-handling model (parallel computation)
## numThread is greater or equal to 2 will call parallel computation
## with given number of threads
fitMRH(dat1, start = c(4, 0.4, 0.1, 1, 0.5), numThreads = 4)
```


We also mention that the sampling functions, `rBMME`, `rMR`, and `rMRH`, are given
for BMME model, moving-resting model and moving-resting-handling model.

##3. Segment dataset analysis

In this section, we are going to show the usage of segment estimation. In contrast
with whole data estimation, the segment estimation only use part of observations.
Segment estimation evaluates the likelihood for each segments of observation and
sums them together as the likelihood be used in MLE process. So, in order to use
segment estimation, we should tell R which part of observations we are going to
use by adding additional 0-n column, where n can be any non-zero integer. The 0's
in this column means discard this observation and other non-zero integers are the
labels for each segments.

```{r, echo = TRUE}
## the name of segment indicator column can be given by user
batch <- c(rep(0, 30), rep(1, 40), rep(0, 20), rep(2, 30), rep(0, 80))
dat2 <- cbind(dat1, batch) ## the result of dat2 is shown in appendix.
```

Then, when we apply `fitBMME`, `fitMR`, `fitMRH` on `dat2`, only segment 1 (batch = 1)
and segment 2 (batch = 2) will be used. Then, we are going to show the code for
fit these three models. (Note that, the name of indicator column should be set
in `segment`.)

```{r, echo=TRUE, eval=FALSE}
## fit BMME model
fitBMME(dat2, start = c(0.5, 0.5), segment = "batch")

## fit moving-resting model with full likelihood
fitMR(dat2, start = c(10, 0.1, 1), segment = "batch", likelihood = "full")

## fit moving-resting model with composite likelihood
fitMR(dat2, start = c(10, 0.1, 1), segment = "batch", likelihood = "composite")

## fit moving-resting-handling model
fitMRH(dat2, start = c(4, 0.4, 0.1, 1, 0.5), segment = "batch", numThreads = 2)
```


##4. Prediction of hidden states

The code for prediction of hidden states is still in process. But you can try and
explore them with `?fitStateMR` and `?fitStateMRH`.










# References
[1] Yan, J., Chen, Y., Lawrence-Apfel, K., Ortega, I. M., Pozdnyakoc, V., Williams, S., and Meyer, T. (2014) A moving-resting process with an embedded Brownian motion for animal movements. Population Ecology. 56(2): 401--415. doi:10.1007/s10144-013-0428-8

[2] Pozdnyakov, V., Elbroch, L., Labarga, A., Meyer, T., and Yan, J. (2017) Discretely observed Brownian motion governed by telegraph process: estimation. Methodology and Computing in Applied Probability. doi:10.1007/s11009-017-9547-6.

[3] Pozdnyakov, V., Elbroch, L., Hu, C., Meyer, T., and Yan, J. (2018+) On estimation for Brownian motion governed by telegraph process with multiple off states. <arXiv:1806.00849>

[4] Pozdnyakov, V. , Meyer, T. , Wang, Y. and Yan, J. (2014), On modeling animal movements using Brownian motion with measurement error. Ecology, 95: 247-253. doi:10.1890/13-0532.1

# Appendix

We show the data for segment estimation here.

```{r, echo=TRUE}
dat2
```


